---
title: "Reference Code for Regressions"
author: "Bohdan Horak"
date: "Tuesday, August 26, 2014"
output: pdf_document
---




### Standard Regression
Getting things started
Using mtcars data set. 

```{r}
library(knitr) ## have to call this before setting global knit options
opts_chunk$set(message=FALSE,results = "asis") 
data(mtcars)
library(xtable) ## <- pretty tables

```

Using mtcars data set. 

```{r}
fit<-lm(mpg~wt,data=mtcars) ## fir is "lm" class
summary(fit)            ## Returns the regression coefficients, 
                        ## hypothesis tests and some residual analysis
```

**Getting Prettier Results**

```{r}
print(xtable(summary(fit)),comment=F) ## Returns a nicer looking 
                                      ## table of the coefficients.  

```



You can also subset the summary to get at more specfic results


```{r}
coef<-summary(fit)$coefficients
residuals<-summary(fit)$coefficents

names(summary(fit))                 ## Full list of names

```

### Behavior of dummy variables

Factor variables will automatically be treated as dummys. The lowest "factor" variable
will automatically be the base case. Example cyl is a factor variable


```{r}
fit2<-lm(mpg~factor(cyl),data=mtcars)
print(xtable(summary(fit2)),comment=F)

```

### Functions of variables - linear transformation/scaling/demeaning

```{r}
fit3<-lm(mpg~I(3+wt*5),data=mtcars) ## notice the capital I
print(xtable(summary(fit3)),comment=F)

```

**or**

```{r}
fit4<-lm(I(mpg*2)~wt,data=mtcars)
print(xtable(summary(fit4)),comment=F)

```

-------


### Anova

```{r}
print(xtable(anova(fit)),comment=F)

```


